A Tale of Two Dictionary Learning Problems

### 2020-08-17 manuscripts/thesis/abstract.tex
final

Two formulations of the dictionary learning problem are considered. 
The first seeks a dictionary for which each signal in a dataset is well-approximated by a linear superposition of only a limited number of dictionary elements.
%he first seeks a dictionary admitting a sparse representation of limited support size for each signal in a dataset. 
%In this case, beyond an original extension of existing guarantees to the noisy regime, a theory of combinatorial designs is introduced to show that the dictionary and corresponding sparse codes are almost always identifiable up to an error commensurate with the noise even if the dictionary fails to satisfy the spark condition, its size is overestimated, or it induces only a polynomial number of distinct sparse supports in the data. 
In this case, existing guarantees are extended to the noisy regime to demonstrate that such dictionaries and the sparse representations they induce are almost always identifiable up to an error commensurate with the noise. Moreover, a theory of combinatorial designs is introduced to demonstrate this is the case even if the dictionary fails to satisfy the spark condition, the data are distributed over only a polynomial set of subspaces spanned by the dictionary, or (to some extent) even if the dictionary is overestimated in size.

The second formulation of the problem seeks a dictionary which minimizes the average number of dictionary elements required to approximate each signal in the dataset. The guarantees in this case are the first of their kind in both in the noiseless and noisy regimes and derive from the fact that, given sufficient data, this second problem actually reduces to an instance of the first. Importantly, in both cases, no constraints whatsoever are imposed on learned dictionaries beyond a natural upper bound on their size.

>>>>

Two formulations of the dictionary learning problem are considered. 
The first seeks a dictionary for which each signal in a dataset is approximated up to some bounded error by a linear superposition of only a limited number of dictionary elements.
%he first seeks a dictionary admitting a sparse representation of limited support size for each signal in a dataset. 
%In this case, beyond an original extension of existing guarantees to the noisy regime, a theory of combinatorial designs is introduced to show that the dictionary and corresponding sparse codes are almost always identifiable up to an error commensurate with the noise even if the dictionary fails to satisfy the spark condition, its size is overestimated, or it induces only a polynomial number of distinct sparse supports in the data. 
In this case, existing guarantees are extended to the noisy regime to show that such dictionaries and the sparse representations they induce are almost always identifiable up to an error commensurate with the approximation error. A theory of combinatorial designs is introduced furthermore to demonstrate this is so even if the dictionary fails to satisfy the spark condition, the data are distributed over only a polynomial set of subspaces spanned by the dictionary, or (to some extent) even if the dictionary is overestimated in size.

The second formulation of the problem seeks a dictionary which minimizes the average number of dictionary elements required to approximate each signal in the dataset up to some bounded error. The guarantees in this case, the first of their kind in both in the noiseless and noisy regimes, derive from the fact that this second problem actually reduces to an instance of the first. Importantly, in both cases, no constraints whatsoever are imposed on learned dictionaries beyond a natural upper bound on their size.

### 2020-08-17 manuscripts/thesis/thesis.tex
final

\begin{dedication}
\null\vfil
%\begin{center} 
I want to dedicate this work to my dad. At the same time, I don't want to, because these results can't be matched to the time with him I traded to produce them, and it feels as though dedicating this work to him would in some sense be measuring his life up against them. The fact is that whatever may be the butterfly effect of this work on the scientific ecosystem, these mathematical truths could have waited, arrived at their own pace, via myself or another, and I could have been more present with my dad as he waited patiently to die of cancer.

I suppose that I couldn't really fathom what it actually meant for this to happen -- to never see or talk to or hug him again -- while the idea of putting original work out there, of crafting my own piece of the overwhelming puzzle that is neuroscience, was tantalizing. Yet it was more-so a feeling of desperation, of having to accomplish something and prove myself worthy of the position I had come to find myself in after 25+ years of schooling and a first failed project of my own design, that I feel largely drove these efforts. Truth be told, I initially had no particular interest in dictionary learning other than as a means to passing my qualifying exam; and by that time, I worried the brain was so dauntingly complex that I should feel lucky to have at least been offered a well-defined problem of arguable relevance to neuroscience to work on, particularly one I felt even remotely equipped to solve myself.

Let's say, then, that this is the first bit of a life's work dedicated to my dad, work devoid of any prestigious position and the associated self-doubt, pressure, or delusions of grandeur; a life dedicated to passionate ideas, to solving problems I care about and that really matter, to loving and caring for others or, as my dad would say, to ``being a mensch". May this dedication serve to hold me publicly accountable to that. Also I love you, mom. I dedicate this and everything to come to you, too.

@@

Finally, a heartfelt thank you to Eric Weiss in particular for introducing me to the student co-ops, and to the Berkeley Student Cooperative (www.bsc.coop) for the work they do providing the amazing experience of living co-operatively in Berkeley. The co-ops changed my life, and I am forever grateful -- thank you so, so much to the hundreds of talented, creative, weird, passionate, intelligent, kind, thoughtful, absurd, inspiring people I have lived with throughout the years for all of the wonderful and weird times. CZ or die. 
\end{acknowledgements}

>>>>

\begin{dedication}
\null\vfil
%\begin{center} 
I want to dedicate this work to my dad. Yet it feels as though that would in some sense be measuring his life up against these results, which can't possibly match the time with him I traded to produce them. The fact is that whatever may be the eventual effect of this work on the scientific ecosystem, these mathematical truths could have waited. They could have arrived at their own pace, via myself or another, and I could have been more present with my dad as he waited patiently to die of cancer.

I suppose that I couldn't really fathom what it actually meant for this to happen -- to never see or talk to or hug him again -- while the idea of putting original work out there, of crafting my own theorems to lay down as pieces of the overwhelming puzzle that is neuroscience, was tantalizing. Yet I feel it was very much a sense of desperation, of having to accomplish something and prove myself worthy of the position I had come to find myself in after 25+ years of schooling and a first failed project, that largely drove these efforts. Truth be told, I initially had no particular interest in dictionary learning other than as a means to passing my qualifying exam; and by that time, I worried the brain was so dauntingly complex that I should feel lucky to have at least been offered a well-defined problem of arguable relevance to neuroscience to work on, particularly one I felt even remotely equipped to solve myself.

So let's just say that this is the first bit of a life's work dedicated to my dad. A life of work uninfluenced by position or status and the associated self-doubt, pressure, or delusions of grandeur; one dedicated to passionate ideas, to solving problems I care about and that really matter, to loving and caring for others or, as my dad would say, to ``being a mensch". May this dedication serve to hold me accountable for that. Also, I love you, mom. I dedicate this and everything to come to you, too.

@@

Finally, a heartfelt thank you to Eric Weiss in particular for introducing me to the student co-ops, and to the Berkeley Student Cooperative (www.bsc.coop) for the work they do providing the amazing experience of living co-operatively in Berkeley. The co-ops changed my life, and I am forever grateful -- thank you so, so much to the hundreds of talented, creative, weird, passionate, intelligent, kind, thoughtful, absurd, and inspiring people I have lived with throughout the years for all of the wonderful and weird times. CZ or die. 
\end{acknowledgements}
